Fast Sorting on Modern Computers

### 2005-08-18 docs/tech_report/quick.tex
minor changes to quicksort predictor stuff on reread

Base quicksort has a 35\% branch misprediction rate, whereas the rate for
memory-tuned quicksort is only 34\%. The main reason for this is the reduction
in branches by not attempting to move the pivot elements.  Attempting to move
these elements usually results in a branch misprediction. The main reason is
that most other elements move at least one position, and thus cause the
insertion sort loop to iterate at least once. But the pivot elements never move
-- they are already in the correct place.

@@

The variation of multi-quicksort that uses sequential search dramatically
increases the number of executed branches, but these branches are highly
predictable. This version actually causes fewer branch mispredictions than
memory-tuned quicksort.

% Theres a lot of branches in the multi quicksort. WE saw in other results that
% the mere instruction cost of this bogs down to algorithm significantly. in the
% binary search version, there are more than twice as many mispredictions,
% despite the fact that there are less than half the branches. (note that i dont
% think this was a wonderful binary search, its heavily biased towards the
% checking the left). binary search predicts correctly 57 percent of the time,
% meaning that it mostly changes direction haphazardly during the search. No
% wonder it performs so badly.

% Note that sequential multi-quicksort has less mispredictions than cache quicksort.
\subsection{Different Medians}

@@

In order to investigate this, we implemented a simple version of quicksort that
uses the middle element as the pivot rather than a median-of-three\footnote{It
is important to note that selecting the pivot based on only a single element can
be particularly inefficient because it removes the natural sentinels at the top
and bottom of the regions to be partitioned that we get from median-of-three. In
Figure \ref{quick-predictors2} we see that this results in an additional branch
to handle the edgecase without a sentinel. Although this branch is almost 100\%
predictable, it is executed a very large number of times.}. The arrays we sort
are random, so the value of this element will also be random. Figure \ref{base
quicksort without a median, with median of 3, and with pseudo-medians of 5, 7
and 9-1} shows the behaviour of each of the important branches in this code. We
see that the \n{i} and \n{j} branches are indeed taken about twice as often as
not-taken, and the \n{i} and \n{j} branches are correctly predicted around 71\%
of the time.

@@

In the preceding paragraphs we have seen that the \n{i} and \n{j} branches are
predictable when we have a pivot that is larger or smaller than the true median
of the array section to be partitioned. An interesting question is the effect of
choosing a better approximation of the true median on branch prediction. Figure
\ref{base quicksort without a median, with median of 3, and with pseudo-medians
of 5, 7 and 9-2} shows the behavior of these branches when median-of-three is
used to select the pivot. Perhaps the biggest effect is that the number of
executions of the \n{i} and \n{j} branches falls by around 15\%. This fall is
partly offset by the comparison branches to compute the median, but these
account for only around 1.25\% of the original branches from the no-median
quicksort. Thus, there is a reduction of almost 14\% in the number of executed
comparison branches. Even more interesting, however, is the bias in the
remaining \n{i} and \n{j} branches. The no-median \n{i} and \n{j} branches are
66\% biased, but the median-of-three branches are only 60\% biased in the taken
direction. The accuracy with which these branches are correctly predicted falls
from 71\% to 64\%.

@@

Figure \ref{quick_median_cycles} shows the cycle count of each of the
quicksorts, showing that a pseudo-median-of-seven offers the greatest trade-off
between the reduced number of branches and instructions, and the increase in
branch mispredictions.

@@

Two areas of current research into quicksort are those of equal keys, and
median-of-greater-than-3. The former affects the algorithmic properties, and
some steps may need to be varied to avoid extra cache misses as a result
of extra copying. The latter simply affects the instruction count, though the
extra comparisons are likely to be mispredicted. These results could be
measured, giving greater insight to this debate.

>>>>

Overall, base quicksort has a 35\% branch misprediction rate, whereas the rate
for memory-tuned quicksort is only 34\%. The main reason for this is the
reduction in branches by not attempting to move the pivot elements.  Attempting
to move these elements usually results in a branch misprediction.  The main
reason is that most other elements move at least one position, and thus cause
the insertion sort loop to iterate at least once. But the pivot elements never
move -- they are already in the correct place.

@@

The variation of multi-quicksort that uses sequential search (see Figure
\ref{base, memory-tuned and multi- quicksorts-4}) dramatically increases the
number of executed branches, but these branches are highly predictable. This
version actually causes fewer branch mispredictions than memory-tuned quicksort.

\subsection{Different Medians}

@@

In order to investigate this, we implemented a simple version of quicksort that
uses the middle element as the pivot rather than a median-of-three\footnote{It
is important to note that selecting the pivot based on only a single element can
be particularly inefficient because it removes the natural sentinels at the top
and bottom of the regions to be partitioned that we get from median-of-three. In
Figure \ref{quick-predictors2} we see that this results in an additional branch
to handle the edgecase without a sentinel. Although this branch is almost 100\%
predictable, it is executed a very large number of times.}. The arrays we sort
are random, so the value of this element will also be random. Figure \ref{base
quicksort without a median, with median of 3, and with pseudo-medians of 5, 7
and 9-1} shows the behaviour of each of the important branches in this code. We
see that the \n{i} and \n{j} branches are indeed taken about twice as often as
not-taken, and the \n{i} and \n{j} branches are correctly predicted
approximately 71\% of the time.

@@

In the preceding paragraphs we have seen that the \n{i} and \n{j} branches are
predictable when we have a pivot that is larger or smaller than the true median
of the array section to be partitioned. An interesting question is the effect of
choosing a better approximation of the true median on branch prediction. Figure
\ref{base quicksort without a median, with median of 3, and with pseudo-medians
of 5, 7 and 9-2} shows the behavior of these branches when median-of-three is
used to select the pivot. Perhaps the biggest effect is that the number of
executions of the \n{i} and \n{j} branches falls by approximately 15\%. This
fall is partly offset by the comparison branches to compute the median, but
these account for only around 1.25\% of the original branches from the no-median
quicksort. Thus, there is a reduction of almost 14\% in the number of executed
comparison branches. Even more interesting, however, is the bias in the
remaining \n{i} and \n{j} branches. The no-median \n{i} and \n{j} branches are
66\% biased, but the median-of-three branches are only 60\% biased in the taken
direction. The accuracy with which these branches are correctly predicted falls
from 71\% to 64\%.

@@

Figure \ref{quick_median_cycles} shows the cycle count of each of the
quicksorts, showing that a median-of-three offers the greatest trade-off
between the reduced number of branches and instructions, and the increase in
branch mispredictions. This shows the cost of the extra branch mispredictions.
It appears that it can be more costly to remove unnecessary entropy than to
endure it.

@@

Two areas of current research into quicksort are those of equal keys, and
median-of-greater-than-three. The former affects the algorithmic properties, and
some steps may need to be varied to avoid extra cache misses as a result
of extra copying. The latter simply affects the instruction count, though the
extra comparisons are likely to be mispredicted. These results could be
measured, giving greater insight to this debate.
